---
title: "Human spike-in experiments"
output:
  workflowr::wflow_html:
      code_folding: hide
---

Human samples are duplicate of 1% spike-in of 8393 (son of Chinese ancestry HG-005) in 8391 (son of Eastern European Ashkenazi Jewish ancestry HG-0020). [Reference](https://www.coriell.org/1/NIGMS/Collections/NIST-Reference-Materials).

```{r setup, include=FALSE}
DOCNAME = "Human mixture duplex data"
knitr::opts_chunk$set(autodep        = TRUE,
                      cache          = FALSE,
                      cache.path     = paste0("cache/", DOCNAME, "/"),
                      cache.comments = FALSE,
                      cache.lazy     = FALSE,
                      echo           = TRUE,
                      error          = FALSE,
                      fig.align      = "center",
                      fig.width      = 7,
                      fig.height     = 5,
                      dev            = c("png"),
                      message        = FALSE,
                      warning        = FALSE)
```

```{r libraries, cache=FALSE, message=FALSE}
library(ggplot2)
library(data.table)
library(dplyr)
library(here)
library(tibble)
library(stringr)
library(Rsamtools)
library(GenomicRanges)
library(seqinr)
library(parallel)
library(readxl)
library(patchwork)
library(RColorBrewer)
library(UpSetR)
library(vcfR)
library(R.utils)
library(knitr)
```

```{r source}
source(here("code/load_data.R"))
source(here("code/plot.R"))
source(here("code/efficiency_nanoseq_functions.R"))
```


```{r paths}
variant_dir <- here("data/human_mixture_vars")
pileup_dir <- here("data/human_mixture_pileup")
region_bed <- here("data/human_mixture_capture_region.bed")
hg002_var_file <- here("data/human_mixture_refs/HG002_GRCh38_1_22_v4.2.1_benchmark.vcf.gz")
hg005_var_file <- here("data/human_mixture_refs/HG005_GRCh38_1_22_v4.2.1_benchmark.vcf.gz")
```


```{r load_variants, cache=FALSE}
sample_names <- c("Human1pR1", "Human1pR2")

# load variant data
var_df <- load_variants(variant_dir, sample_names)
hg002_vars <- read.vcfR(hg002_var_file, verbose = FALSE)
hg005_vars <- read.vcfR(hg005_var_file, verbose = FALSE)

hg002v <- data.frame(hg002_vars@fix)
hg005v <- data.frame(hg005_vars@fix)

# get capture regions
regions <- read.delim(region_bed, sep = "\t", header = FALSE)
grx <- GRanges(seqnames = regions$V1,
               ranges = IRanges(start = regions$V2, end = regions$V3))
```

## Variant Upset plot

Here we remove any "N" variant calls and INDELs and compare the overlaps for on- and off-target variant calls.

```{r variant_filtering, fig.width = 9, fig.height = 6}
# remove any N calls and INDELs
var_df <- filter(var_df, ALT != "N") %>%
            filter(., (REF %>% str_split("") %>% lapply(., length) %>% unlist) == 1) %>%
            mutate(ALT = lapply(ALT, filter_out_indels) %>% as.character()) %>%
            filter(., (ALT %>% str_split("") %>% lapply(., length) %>% unlist) == 1)

# filter out off-target reads
vrx <- GRanges(seqnames = var_df$CHROM,
               ranges = IRanges(start = as.numeric(var_df$POS),
                                end = as.numeric(var_df$POS) + 1))
var_df$on_target <- overlapsAny(vrx, grx)

# calculate vafs
alt_dep <- apply(var_df, 1, get_alt_dep_nvc) %>% t() %>% data.frame()
var_df$AC <- alt_dep$X1
var_df$DP <- alt_dep$X2
var_df$VAF <- var_df$AC / var_df$DP

# make upsetplot
ulist <- NULL
for(sample in sample_names) {
    ont_ids <- var_df[var_df$sample %in% sample & var_df$on_target,]$id
    oft_ids <- var_df[var_df$sample %in% sample & !var_df$on_target,]$id
    ulist[[paste(sample, "on_target")]] <- ont_ids
    ulist[[paste(sample, "off_target")]] <- oft_ids
    
}

upset(fromList(ulist), order.by="freq", nsets=4)
```

Checking on-target rate from the bam files (based on reads that fall within the region using `samtools view -c -L <region_bed> <consensus_bam>`) yields a higher on-target rate than the variant analysis would suggest:

    
    	            Human1pR1	Human1pR2
    capture_region	188409	    122466
    total_reads	    330368	    233768
    on_target	    0.5703	    0.5239


## Variant allele frequencies

Here we plot the allelic frequencies per-replicate in three plots:

 - all filtered variant calls: no frequency or target filtering
 - on-target variant calls: only on-target variant calls (in capture region)
 - VAF-filtered on-target variant calls: all on-target variants under <0.3\% VAF

```{r vaf_plots, fig.width = 9, fig.height = 4}
ggplot(var_df, aes(VAF)) +
    geom_histogram(binwidth = 0.05) +
    facet_grid(~sample) +
    theme_minimal() +
    ggtitle("All filtered variant calls")

ggplot(var_df[var_df$on_target,], aes(VAF)) +
    geom_histogram(binwidth = 0.05) +
    facet_grid(~sample) +
    theme_minimal() +
    ggtitle("On-target variant calls")

ggplot(var_df[var_df$on_target & var_df$VAF < 0.3,], aes(VAF)) +
    geom_histogram(binwidth = 0.01) +
    facet_grid(~sample) +
    theme_minimal() +
    ggtitle("On-target variant calls < 0.3 VAF")
```

## Variant comparison

Given the referene information, we check how many SNPs are present in the capture area. We also filter out any variants that are common between the two samples (if the same variant appears in both samples, we can"t differentiate the calls without some kind of phasing).

```{r variant_comparison}
# filter out any INDELs
hg005v <- filter(hg005v, (ALT %>% str_split("") %>% lapply(., length) %>% unlist) == 1) %>%
          filter(., (REF %>% str_split("") %>% lapply(., length) %>% unlist) == 1)

# construct Granges for hg005 SNPs and keep only SNPs in capture area
hg5x <- GRanges(seqnames = hg005v$CHROM,
                ranges = IRanges(start = as.numeric(hg005v$POS),
                                 end = as.numeric(hg005v$POS) + 1),
                variant = hg005v$ALT)
hg5x <- hg5x[overlapsAny(hg5x, grx) %>% suppressWarnings()]

# construct Granges for hg002
hg2x <- GRanges(seqnames = hg002v$CHROM,
                ranges = IRanges(start = as.numeric(hg002v$POS),
                                 end = as.numeric(hg002v$POS) + 1),
                variant = hg002v$ALT)

# check variants that overlap, we will keep these if they call a different base
unique_vars <- hg5x[overlapsAny(hg5x, hg2x)]$variant != hg2x[overlapsAny(hg2x, hg5x)]$variant
hg5x <- c(hg5x[!overlapsAny(hg5x, hg2x)], hg5x[overlapsAny(hg5x, hg2x)][unique_vars])

kable(hg5x)
kable(var_df[var_df$POS %in% start(hg5x) & var_df$CHROM %in% seqnames(hg5x),])
```

As a sanity check, bcftools v1.17 mpileup was run on the bam files no BAQ and min base Q of 0. Below are the VAFs for the consensus bam variants:

```{r check_cons_pileup}
h1mp_cons <- read.vcfR(file.path(pileup_dir, "Human1pR1_pileup.vcf"),
                      verbose = FALSE)
h2mp_cons <- read.vcfR(file.path(pileup_dir, "Human1pR2_pileup.vcf"),
                      verbose = FALSE)

h1mp_cons <- data.frame(h1mp_cons@fix) %>% cbind(., data.frame(h1mp_cons@gt))
h2mp_cons <- data.frame(h2mp_cons@fix) %>% cbind(., data.frame(h2mp_cons@gt))
h1mp_cons$ID <- paste(h1mp_cons$CHROM, h1mp_cons$POS, sep = "_")
h2mp_cons$ID <- paste(h2mp_cons$CHROM, h2mp_cons$POS, sep = "_")

hg5x$ID <- paste(seqnames(hg5x), start(hg5x), sep = "_")
h1mp_cons <- h1mp_cons[h1mp_cons$ID %in% hg5x$ID,] %>% distinct()
h2mp_cons <- h2mp_cons[h2mp_cons$ID %in% hg5x$ID,] %>% distinct()

h1mp_cons$VAF <- apply(h1mp_cons, 1, get_variant_fraction, hg5x)
h2mp_cons$VAF <- apply(h2mp_cons, 1, get_variant_fraction, hg5x)

kable(h1mp_cons)
kable(h2mp_cons)
```

Below are the VAFs for the raw bam variants:

```{r check_raw_pileup}
h1mp_raw <- read.vcfR(file.path(pileup_dir, "raw/Human1pR1_pileup.vcf"),
                      verbose = FALSE)
h2mp_raw <- read.vcfR(file.path(pileup_dir, "raw/Human1pR2_pileup.vcf"),
                      verbose = FALSE)

h1mp_raw <- data.frame(h1mp_raw@fix) %>% cbind(., data.frame(h1mp_raw@gt))
h2mp_raw <- data.frame(h2mp_raw@fix) %>% cbind(., data.frame(h2mp_raw@gt))
h1mp_raw$ID <- paste(h1mp_raw$CHROM, h1mp_raw$POS, sep = "_")
h2mp_raw$ID <- paste(h2mp_raw$CHROM, h2mp_raw$POS, sep = "_")

h1mp_raw <- h1mp_raw[h1mp_raw$ID %in% hg5x$ID,] %>% distinct()
h2mp_raw <- h2mp_raw[h2mp_raw$ID %in% hg5x$ID,] %>% distinct()

h1mp_raw$VAF <- apply(h1mp_raw, 1, get_variant_fraction, hg5x)
h2mp_raw$VAF <- apply(h2mp_raw, 1, get_variant_fraction, hg5x)

kable(h1mp_raw)
kable(h2mp_raw)
```

