---
title: "Rare mutation detection model"
output:
  workflowr::wflow_html:
      code_folding: hide
---

```{r knitr, include = FALSE}
DOCNAME = "Simulation_Benchmarking"
knitr::opts_chunk$set(autodep        = TRUE,
                      cache          = FALSE,
                      cache.path     = paste0("cache/", DOCNAME, "/"),
                      cache.comments = FALSE,
                      echo           = TRUE,
                      error          = FALSE,
                      fig.align      = "center",
                      fig.width      = 9,
                      fig.height     = 5,
                      dev            = c("png"),
                      message        = FALSE,
                      warning        = FALSE)
```

```{r libraries, cache = FALSE}
library(ggplot2)
```

# Model

Q: what is the lowest % VAF of mutation we can reliably detect (at >95% confidence) using Nanoseq on bulk WES?

## Assumptions

* We consider a human diploid genome (6.4Gb) without SCNAs
* The mutation is heterozygous
* We sequence 15,000 cells
* Our duplex rate is optimal ~81%
* Our ligation efficiency is 20%
* We assume a duplex mutation call equals a real mutation (theoretical error rate of duplex sequencing is <10^-9)

## Probability of sequencing a mutant cell

Let $f$ be the probability of sequencing a mutation from a single fragment, on both strands.

$f = (v / p) d.l$

Where:

* $v$ = target VAF
* $p$ = ploidy (2)
* $d$ = duplex efficiency (0.81)
* $l$ = ligation efficiency

We assume the probability of selecting a mutant cell is binomially distributed. We want to know the probability of selecting at least one mutant cell:

$P(Bin(f, m)) > 0)$ = 0.95

This is equivalent to:

$P(Bin(f, m)) = 0)$ = 0.05

Where $m$ is the number of mutant cells cells ($15000 (v . 2)$).

For a range of possible VAFs, $V = \{0.001, 0.002 .. 0.05\}$, we can plot the probability of missing the mutant cell.

```{r vaf_plot}
d = 0.81
l = 0.2
v = seq(0.001, 0.05, 0.001)
f = (v / 2) * d * l
n = 15000
m = n * (v * 2)

vafs <- data.frame(vaf=v,
                   p=dbinom(0, m, v))

ggplot(vafs, aes(vaf, p)) +
  geom_point() +
  theme_bw() +
  geom_hline(yintercept=0.05, alpha=0.4)
```
For this range of VAFs, 0.01 (1%) is the smallest VAF for which the probability of missing the mutant is approximately 0.05.

```{r}
deviation <- abs(0.05 - vafs$p)
print(vafs[which(deviation == min(deviation)),])
```

